---
title: "Life is short"
author: "Göran Broström"
date: "April 11, 2018"
output: html_document
---

```{r datain}
library(skum)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

This is a preparation for a talk with the title

"life is short but unlimited"

that will be given (for the first time) at CEDAR, Umeå University, on the 17th of May, 2018, at 1300 hours. Inspired by the paper


## High ages at death, Skellefteå and Umeå 1801--1950

```{r maxyearskum}
years <- 1801:1950
ages <- numeric(length(years))
dobs <- obs[obs$sluttyp == 2, c("sex", "birthdate", "exit", "region")]
dobs$deathdate <- dobs$birthdate + dobs$exit
dobs$event <- 1
for (i in seq_along(years)){
    nu <- dobs[floor(dobs$deathdate) == years[i], "exit"]
    ages[i] <- max(nu)
}
plot(years, ages, type = "p", col = "blue", xlab = "Year", ylab = "Max death age") 
abline(h = 100, col = "red")
```

## The Human Mortality Database

```{r dataswed}
deaths <- read.table("data/HMD/Deaths_1x1.txt", header = TRUE, skip = 3, stringsAsFactors = FALSE, na.strings = ".")
##head(deaths)
expos <- read.table("data/HMD/Exposures_1x1.txt", header = TRUE, skip = 3, stringsAsFactors = FALSE, na.strings = ".")
##head(expos)
rates <- read.table("data/HMD/Mx_1x1.txt", header = TRUE, skip = 3, stringsAsFactors = FALSE, na.strings = ".")
str(rates)
```

We note that the variable *Age* is a *character variable*: It is because the last, open class has the label "110+". We change that to "110" and make *Age* a numeric variable.

```{r makeAge_numeric}
deaths$Age[deaths$Age == "110+"] <- "110"
deaths$Age <- as.integer(deaths$Age)
##
expos$Age[expos$Age == "110+"] <- "110"
expos$Age <- as.integer(expos$Age)
##
rates$Age[rates$Age == "110+"] <- "110"
rates$Age <- as.integer(rates$Age)
```

### Females

```{r females}
females <- rates[["Female"]]
fem <- matrix(females, nrow = 111)
colnames(fem) <- 1751:2016
rownames(fem) <- 0:110
```

### Calendar time averages 

The table with ten-year agerages is loaded.

```{r averrate}
aver <- read.table("data/HMD/Mx_1x10.txt", header = TRUE, skip = 3, stringsAsFactors = FALSE, na.strings = ".")
afemales <- aver[["Female"]]
afem <- matrix(afemales, nrow = 111)
rownames(fem) <- 0:110
```

## The International Database on Longevity

```{r idldata}
usa <- read.table("data/IDL/usa.csv", header = TRUE, sep = ";")
x <- with(usa, tapply(Age.days. / 365.25, Date.of.death, max))
plot(1980:2003, x, xlab = "Year", ylab = "Age", ylim = c(110, 120), col = "blue")
abline(h = 110)
abline(h = 111:120, lty = 3, col = "green")    
with(usa, points(Date.of.death, Age.days. / 365.25, cex = 0.25, col = "blue"))
```