---
title: "fromFobLisa, Version RDS"
author: "Göran Broström"
date: "May 28, 2018"
output:
  html_document:
    df_print: paged
  pdf_document: default
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE,cache.lazy = FALSE)
options(stringsAsFactors = FALSE)
library(stringr)
##library(tidyr)
library(dplyr)
```

# Introduction

**NOTE**: This version aims at histories above 90 years of age. Selection in all *Fob*s: All above age 85 (exact).

```{r setAgeInterval}
ageIvl <- c(90, 115) ## Noone live beyond age 115
followUp <- 1 # Look ahead period years, we pick frpm ageIvl[1] - followUp to ageIvl[2]
```

We used the
knowledge that the yearly *lisa* data relates to *December 31* the given
year, while the *FoB* data relate to the date *November 1* the given
year. These facts have the following implications for our plan: For *FoB*
data, we follow individuals from their appearance in a FoB until the next
FoB five years later, even if they do not appear there. The exception is if
they die within five years after the first appearance. So, *fob60* induces
a time interval *from November 1, 1960 to October 31, 1965*, and so on. 
Since we do not have birth dates but only birth year and quarter,
we decide to use as birthdate the *midpoint of the birth quarter*. 
We work with all ages 0-110, and only as the final step, `age.window` is called.

The following FoBs are treated similarly, until we reach the Fob
in 1985. Since the first *lisa* date is December 31, 1986, this "fob
interval" starts at November 1, 1985, and ends at December 31, 1986, that
is, it is 14 months long. 

New version: 'putonssyk' changed. (24 May 2018)

# Fixed data

We start with the *inddata* file and treat it as a *fixed-data*
file (no time-varying info). 
We find the birth dates (see above), death dates (for those who are observed die), and the "date last seen".

Note that 37 individuals in *inddata* have no information at all, so they are removed.

```{r frominddata, cache = TRUE}
inddata <- readRDS("../Data/U18006_RDS/inddata.rds")
inddata <- inddata[inddata$FODELSEAR > 1850, ]
inddata <- inddata[inddata$FODELSEAR < 2015 - ageIvl[1] + followUp + 1, ] # So reach 'ageIvl[1]' before 2015-01-01.
per <- inddata[!is.na(inddata$KON) & !is.na(inddata$FODELSEAR), ] # Not necessary ...
per$birthdate <- per$FODELSEAR + 0.125  +  (per$FODELSEKVARTAL - 1) / 4
rm(inddata)
```

Now we put on *death date*, *death cause*, and *KOMMUN of death*, from *socstyr*, on *per*:

```{r styronper, cache = TRUE}
dors <- readRDS("../Data/U18006_RDS/dors.rds")
indx <- match(per$LINNEID, dors$LINNEID)
per$DODSDAT <- dors$DODSDAT[indx]
fixd <- !is.na(per$DODSDAT) & (per$DODSDAT < 1960010) # Only month and year
per$DODSDAT[fixd] <- per$DODSDAT[fixd] * 100 + 15
per$deathdate <- as.numeric(skum::tillTid(per$DODSDAT))
per$dodsKOMMUN <- dors$KOMMUN[indx]
per$dodsALDER <- dors$ALDER[indx]
per$ULORSAK <- dors$ULORSAK[indx]
rm(dors)
```

# The Lisa data

The *lisa* data is slightly simpler because it is yearly with cutoff dates 12-31/01-01 each new-year. It starts off at 1987-01-01 (lisa1986) and ends at 2013-12-31 (lisa2013). So the time interval a *lisa* covers is *AR + 1*! But note that the *socstyr* data now cover deaths *before January 1, 2015, so *lisa2013* is now in effect useful to us.  

```{r lisadat}
lisa <- readRDS("../Data/U18006_RDS/lisa.rds")[, c("LINNEID", "AR", "KOMMUN", "CIV")]
##lisa <- lisa[(lisa$KOMMUN %in% ourkom) & (lisa$LINNEID %in% per$LINNEID), ]
lisa <- lisa[(lisa$LINNEID %in% per$LINNEID), ]
ar <- 1986:2013
days <- as.numeric(skum::tillTid(paste(ar, "1231", sep = "")))
```

Get *birthdate* and *deathdate* from *per*:

```{r getbirthdate}
indx <- match(lisa$LINNEID, per$LINNEID)

lisa$birthdate <- per$birthdate[indx]
lisa$deathdate <- per$deathdate[indx]
```

Create individual Lisas (1986, 1987, ..., 2013)

```{r rawlisa, cache = FALSE}
source("R/extractLisa.R")
n <- length(ar)
##fobbar <- vector(mode = "list", length = n)
##names(fobbar) <- paste("fob", (ar %% 100), sep = "")
dors <- readRDS("../Data/U18006_RDS/dors.rds")

for (i in 1:n){
    lisadata <- extractLisa(ar[i], followUp = 1, data = lisa)
    ##fobbar[[n]] <- extractFob(ar[n], followUp = 11 + 61 / 365, data = fob)
    ##fobdata <- do.call(rbind, fobbar)
    ###rm(fobbar)
    lisadata <- lisadata[order(lisadata$LINNEID, lisadata$enter), ]
    ##
    rownames(lisadata) <- 1:NROW(lisadata)

    indx <- match(lisadata$LINNEID, dors$LINNEID)
    lisadata$ULORSAK <- str_trim(as.character(dors$ULORSAK[indx]))
    ##putONssyk3
    ##source("R/ssyk3TOhisclass.R")
    ##ssyk3TOhisclass
    ##lisadata$hisclass <- ssyk3TOhisclass(lisadata$SSYK3)
    ##is.na(lisadata$hisclass) <- with(lisadata, is.na(hisclass) |
      ##                                 (!is.na(hisclass) & hisclass == -1))
    ##lisadata$id <- lisadata$LINNEID
    ##load("../../Data/sex.rda")

    indx <- match(lisadata$LINNEID, per$LINNEID)
    lisadata$KON <- per$KON[indx]
    ##fobb$urban <- fobb$FORSAMLING %in% c(248001, 248002, 284004, 284005,  ## This must be fixed!
      ##                                   248011, 248201, 248204)
    civst <- numeric(NROW(lisadata))
    civst[lisadata$CIV %in% 1] <- 1 # Ogift
    civst[lisadata$CIV %in% c(2, 5)] <- 2 # Gift, sambo
    civst[lisadata$CIV %in% c(3, 4, 6, 7)] <- 3 # skild, änka
    is.na(civst) <- civst == 0
    lisadata$civst <- civst
    ##fobb$civst <- factor(civst,
      ##                   labels = c("unmarried", "married", "prev.married"))
    lisadata$sex <- lisadata$KON
    ##levels(fobb$sex) <- c("male", "female")
    lisadata <- lisadata[, c("LINNEID", "AR", "KOMMUN", "birthdate", "sex",
                             "enter", "exit", "event",
                             "civst", "ULORSAK")]
    lisadata <- eha::age.window(lisadata, ageIvl)
    saveRDS(lisadata, file = paste("data/LISA/lisa", ar[i], ".rds", sep = ""))
    rm(lisadata)
}
```

