---
output:
  beamer_presentation:
    keep_tex: true  
    fig_caption: no
    fig_width: 5
    includes:
      before_body: gb_beforebody.txt
      in_header: gb_header.txt
  ioslides_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.width = 11, warning = FALSE)
```

## Extremes, December 2017

\includegraphics[width=\textwidth]{myfigs/rootzen.png}

Received from: [Zentralblatt für Mathematik](https://zbmath.org).

## My review

\includegraphics[width=\textwidth]{myfigs/zbmath.png}

## What it is about

*   Is human life \textcolor{blue}{unlimited}?
*   Are there any \textcolor{blue}{differences in survival} at extreme age between 
    *   women and men, 
    *   different socioeconomic groups, 
    *   etc?

<!--
*   What will the record life-length be the coming 25 years?    
*   Etc.

But **most important**: **Bad statistics** in previous research.
-->

## Broken Limits to Life Expectancy: Science 2002

\includegraphics[width=0.9\textwidth]{myfigs/oeppenvaupel.png}

\scriptsize

Authors: James Vaupel and Jim Oeppen. \alert{Gain / year: } 3 months.

## Sweden 1968-2014: Remaining life at birth

```{r gainperyear, eval = TRUE}
library(Sweden)
library(eha)
##source("R/mpch.R")
fd <- getDeaths(sex = "females")
fp <- getPop(sex = "females", years = 1968:2017)
frat <- fd / fp
md <- getDeaths(sex = "males")
mp <- getPop(sex = "males", years = 1968:2017)
mrat <- md / mp
cut <- 1:100
favg <- numeric(50)
mavg <- numeric(50)
for (i in 1:50){
    favg[i] <- mpch(cuts = cut, levels = frat[, i] )
    mavg[i] <- mpch(cuts = cut, levels = mrat[, i] )    
}
year <- 1968:2017
plot(year, favg, type = "b", col = "red", ylim = c(70, 85), axes = FALSE, xlab = "Year", ylab = "Age")
axis(1, at = c(seq(1968, 2008, by = 10), 2017))
axis(2)
box()
lines(year, mavg, type = "b", col = "blue")
text(1980, 82, "Women (1.83 months / year)", col = "red")
text(2000, 73.2, "Men (2.44 months / year)", col = "blue")
abline(h = c(70, 75, 80, 85), lty = 3)
```

## Sweden 1968-2014: Remaining life at age 90

```{r gainperyear90, eval = TRUE}
library(Sweden)
fd <- getDeaths(sex = "females", ages = 90:100)
fp <- getPop(sex = "females", years = 1968:2017, ages = 90:100)
frat <- fd / fp
md <- getDeaths(sex = "males", ages = 90:100)
mp <- getPop(sex = "males", years = 1968:2017, ages = 90:100)
mrat <- md / mp
cut <- 1:10
favg <- numeric(50)
mavg <- numeric(50)
for (i in 1:50){
    favg[i] <- mpch(cuts = cut, levels = frat[, i] )
    mavg[i] <- mpch(cuts = cut, levels = mrat[, i] )    
}
year <- 1968:2017
plot(year, favg + 90, type = "b", col = "red", ylim = c(90, 95), axes = FALSE, xlab = "Year", ylab = "Age")
axis(1, at = c(seq(1968, 2008, by = 10), 2017))
axis(2)
box()
lines(year, mavg + 90, type = "b", col = "blue")
text(1980, 94.4, "Women (7 days / year)", col = "red")
text(2000, 92.6, "Men (6 days / year)", col = "blue")
abline(h = 90)
abline(h = c(93, 94, 95), lty = 3)
```

## Sweden 1969-2013: Maximum life

```{r maxlife}
s90 <- readRDS("mydata/per_90.rds")
s90$sex = factor(s90$KON, labels = c("man", "woman"))
s90$sex <- relevel(s90$sex, ref = "woman")
s90 <- eha::cal.window(s90, c(1969, 2014))
s90 <- s90[, c("enter", "exit", "event", "birthdate", "sex")]
i <- 0
fm <- numeric(45)
mm <- numeric(45)
for (year in 1969: 2013){
    i <- i + 1
    xx <- eha::cal.window(s90, c(year, year + 1))
    mm[i] <- max(xx$exit[xx$sex == "man"])
    fm[i] <- max(xx$exit[xx$sex == "woman"])
}
plot(1969:2013, fm, type = "b", col = "red", ylim = c(100, 115), xlab = "Year", ylab = "Age", axes = FALSE, lww = 1.5)
axis(1, at = c(1969, 1979, 1989, 1999, 2013))
axis(2, las = 1)
axis(4, las = 1)
abline(h = c(100, 105, 110, 115), lty = 3)
box()
lines(1969:2013, mm, type = "b", col = "blue")
text(1981, 112, "Women", col = "red")
text(1990, 106, "Men", col = "blue")
```

## Chiyo Miyako, Japan

\includegraphics[width=\textwidth]{myfigs/ChiyoMiyako2.jpg}

World's oldest person since 21 April, 2018. Age: 117.

(Age on picture: 114.)

## Aftonbladet, 27 juli 2018

\includegraphics[width=\textwidth]{myfigs/chiyo_dead.png}

**Now oldest:** \textcolor{blue}{Kane Tanaka}, Japan (115 years and 205 days).

## SCB data on old age mortality

Deaths by age and year

```{r scb}
##library(Sweden)
x <- Sweden::getDeaths(ages = 97:100, years = 2007:2014)
rownames(x) = c(97:99, "100+")
knitr::kable(x)         
```

\textcolor{red}{Not} very useful (100+).

## International Database on Longevity

[https://www.supercentenarians.org](https://www.supercentenarians.org)

## Gerontology Research Group database

[http://www.grg.org](http://www.grg.org)

\includegraphics[width=\textwidth]{myfigs/grg.png}

## The Human Mortality Database

[https://www.mortality.org](https://www.mortality.org)

*   Contains no individual data.
*   But very \textcolor{blue}{useful tables}.

## svt.se, October 2016

\begin{center}
\includegraphics[width = \textwidth]{myfigs/svt3.png}
\end{center}
## Jeanne Calment, age 122.

\includegraphics[width = \textwidth]{myfigs/NYT.png}

## Nature, 2016

\includegraphics[width=\textwidth]{myfigs/nat16.png}

## Dong, Milholland & Vijg, Nature 2016.

\begin{columns}
\begin{column}{0.5\textwidth}
\includegraphics[width=\textwidth, height=0.7\textheight]{myfigs/repage.png}
\end{column}
\begin{column}{0.45\textwidth}
\scriptsize
\begin{itemize}
\item International Database on Longevity
\item Highest age at death each year
\item Linear regression with breakpoint
\item Increasing trend broken!
\end{itemize}

\end{column}
\end{columns}

\begin{center}
Any \textcolor{red}{problems} with this "statistical analysis"?
\end{center}

## Dong, Milholland & Vijg, Nature 2016: Problems.

\begin{columns}
\begin{column}{0.5\textwidth}
\includegraphics[width=\textwidth, height=0.7\textheight]{myfigs/repage.png}
\end{column}
\begin{column}{0.45\textwidth}
\scriptsize
\begin{itemize}
\item International Database on Longevity
\item Highest age at death each year
\item Linear regression with breakpoint
\item Increasing trend broken!
\end{itemize}

\end{column}
\end{columns}

\scriptsize

*   Ad hoc choice of cutpoint.
*   Very short time series to confirm a trend break.
*   What about the survivors?
*   Data are extreme values: full data?

## The full data

```{r readidl, fig.width = 13, fig.height = 6}
par(cex = 1.5)
usa <- read.table("data/IDL/USA/usa.csv", header = TRUE, sep = ";")
usa$year <- usa$Date.of.death
usa$age <- usa$Age.days. / 365.2425
usa$date <- usa$year + runif(NROW(usa), 0, 0.999)
with(usa, plot(date, age, pch = ".", cex = 5, las = 1, xlim = c(1968, 2008), ylim = c(110, 123), 
     xlab = "Date", ylab = "Age", axes = FALSE))
axis(1, at = c(1968, 1980, 1987, 1996, 2004, 2007), las = 2)
axis(2, las = 1)
box()
##
uk <- read.table("data/IDL/UK/uk.csv", header = TRUE, sep = ";")
uk$age <- uk$Age.days. / 365.2425
uk$year <- as.numeric(substr(uk$Date.of.death, 7, 10))
uk$date <- uk$year + runif(NROW(uk), 0, 0.999)
points(uk$date, uk$age, col = "red", pch = ".", cex = 5)
##
france <- read.table("data/IDL/France/france.csv", header = TRUE, sep = ";")
france$age <- france$Age.days. / 365.2425
france$year <- as.numeric(substr(france$Date.of.death, 7, 10))
france$date <- france$year + runif(NROW(france), 0, 0.999)
points(france$date, france$age, col = "blue", pch = ".", cex = 5)
japan <- read.table("data/IDL/Japan/japan.csv", header = TRUE, sep = ";")
japan$age <- japan$Age.days. / 365.2425
japan$year <- as.numeric(substr(japan$Date.of.death, 7, 10))
japan$date <- japan$year + runif(NROW(japan), 0, 0.999)
points(japan$date, japan$age, col = "orange", pch = ".", cex = 5)
abline(v = 1968:2007)
text(2003, 122.5, "< Jeanne Calment", col = "blue", cex = 1 / 1.5)
jeanne <- france[france$age > 120, c("age", "date")]
up <- jeanne$age - 110
x <- c(jeanne$date - up, jeanne$date)
y <- c(jeanne$age - up, jeanne$age)
lines(x, y, lty = 3)
```

\scriptsize

\begin{tabular}{l|rr}
Country & First & Last \\ \hline
UK      & 1968 & 2006 \\
USA     & 1980 & 2003 \\
France  & 1987 & 2003  \\
Japan & 1996 & 2005 \\
\end{tabular}


## Number of deaths by year

```{r deathsbyyear}
par(cex = 1)
vars <- c("year", "age", "date")
all <- rbind(uk[, vars],
             usa[, vars],
             france[, vars],
             japan[, vars])
year <- seq(min(all$year), max(all$year))
deaths <- numeric(length(year))
for (i in 1:length(deaths)){
    deaths[i] <- sum(all$year == year[i])
}
barplot(deaths, names.arg = year, xlab = "Year", 
        ylab = "No. of deaths", las = 2)
abline(h = 0)
```

## Nature, 2017

\includegraphics[width=\textwidth]{myfigs/nat17.png}

## Nature, 2017

\begin{columns}

\begin{column}{0.46\textwidth}

\includegraphics[width=\textwidth]{myfigs/discnat1.png}

\end{column}

\begin{column}{0.46\textwidth}

\includegraphics[width=\textwidth]{myfigs/discnat2.png}

\end{column}

\end{columns}

<!--
## Nature 2017

\includegraphics[width=\textwidth]{myfigs/discnat1.png}

-->




<!--
## The Human Mortality Database

```{r dataswed, echo=FALSE}
deaths <- read.table("data/HMD/Deaths_1x1.txt", header = TRUE, skip = 3, stringsAsFactors = FALSE, na.strings = ".")
##head(deaths)
expos <- read.table("data/HMD/Exposures_1x1.txt", header = TRUE, skip = 3, stringsAsFactors = FALSE, na.strings = ".")
##head(expos)
rates <- read.table("data/HMD/Mx_1x1.txt", header = TRUE, skip = 3, stringsAsFactors = FALSE, na.strings = ".")
str(rates)
```

## The International Database on Longevity

```{r idldata, fig.width=10}
library(lubridate)
options(stringsAsFactors = FALSE)
usa <- read.table("data/IDL/USA/usa.csv", header = TRUE, sep = ";")
usa$year <- usa$Date.of.death
usa$age <- usa$Age.days. / 365.25
usa$country = "usa"
uk <- read.table("data/IDL/UK/uk.csv", header = TRUE, sep = ";")
uk$age <- uk$Age.days. / 365.25
uk$year <- year(as.Date(uk$Date.of.death, format = "%d/%m/%Y"))
uk$country <- "uk"
japan <- read.table("data/IDL/Japan/japan.csv", header = TRUE, sep = ";")
japan$year <- year(as.Date(japan$Date.of.death, format = "%d/%m/%Y"))
japan$age <- japan$Age.days. / 365.25
japan$country <- "japan"
france <- read.table("data/IDL/France/france.csv", header = TRUE, sep = ";")
france$year <- year(as.Date(france$Date.of.death, format = "%d/%m/%Y"))
france$age <- france$Age.days. / 365.25
france$country <- "france"
all <- rbind(usa[, c("year", "age", "Sex", "country")],
             uk[, c("year", "age", "Sex", "country")],
             france[, c("year", "age", "Sex", "country")],
             japan[, c("year", "age", "Sex", "country")])
```

```{r plotthem}
x <- with(all, tapply(age, year, max))
years <- sort(unique(all$year))
plot(years, x, xlab = "Year", ylab = "Age", ylim = c(110, 123), col = "blue", axes = FALSE)
axis(1, at = c(1968, 1980, 1990, 2003, 2006))
axis(2, las = 1)
box()
abline(h = 110)
abline(h = 111:122, lty = 3, col = "green")    
abline(v = c(1968, 2006))
with(all, points(year, age, cex = 0.25, col = "blue"))
with(all[all$country =="usa", ], points(jitter(year, amount = 0.25), age, cex = 0.25, col = "red"))
```


## Swedish data, 110+

```{r sweddat}
s105 <- readRDS("mydata/per_105.rds")
s110 <- eha::age.window(s105, c(110, 120))
##with(s110[s110$event == 1, ], plot(firstSeen, firstSeen -  birthdate))
dead <- s110[s110$event == 1, ]
x1 <- dead$firstSeen
x2 <- dead$deathdate
dur <- x2 - x1
y2 <- dead$exit
y1 <- y2 - dur
plot(c(x1[1], x2[1]), c(y1[1], y2[1]), type = "l", xlim = c(1969, 2015), ylim = c(105, 115),
     xlab = "Date", ylab = "Age")
for (i in 2:NROW(dead)){
    lines(c(x1[i], x2[i]), c(y1[i], y2[i]))
}
cens <- s110[s110$event == 0, ]
x1 <- cens$firstSeen
x2 <- cens$lastSeen
dur <- x2 - x1
y2 <- cens$exit
y1 <- y2 - dur
for (i in 1:NROW(cens)){
    lines(c(x1[i], x2[i]), c(y1[i], y2[i]), col = "red")
}
males <- cens[cens$KON == 1, ]
points(males$lastSeen, males$exit, col = "red")
males <- dead[dead$KON == 1, ]
points(males$deathdate, males$exit, col = "black")
abline(h = 110)
abline(v = c(1969, 2014, 2015), lty = 2)
```
-->

## Rootzén and Zholud

Two serious blows:

1.   \textcolor{red}{Extreme values} do \textcolor{blue}{not} behave like "normal" data.
2.   The data suffer from \textcolor{red}{length bias}.

and one conclusion:

1.   There is \textcolor{red}{no aging} after age 110!

## The Generalized Pareto (GP) distribution

\begin{equation*}
h(x; (\gamma, \sigma)) = \frac{1}{\sigma (1 + \gamma x / \sigma)_{+}}
\end{equation*}

```{r gphaz, fig.width = 11, fig.height = 4}
oldpar <- par(mfrow = c(1, 3), lwd = 1.5, cex.main = 2)
h <- function(x){
    den <- 1 + gamma * x / sigma
    den[den <= 0] <- 0
    1 / (sigma * den)
}
gamma <- -0.25
sigma = 1.3
x <- seq(0, 10, length = 1000)
y1 <- h(x)
plot(x, y1, type = "l", ylim = c(0, 5), col = "blue", axes = FALSE, xlab = "Age", 
     ylab = "Force of mortality", main = expression(paste("(a) ", gamma,  " = -0.25")))
abline(v = - sigma / gamma, lty = 3)
axis(1, at = c(0, 5, 10), labels = c(110, 115, 120))
axis(2, las = 1)
box()
abline(h = 0)
gamma = 0
y2 <- h(x)
plot(x, y2, type = "l", ylim = c(0, max(y2) * 2), col = "blue", axes = FALSE, xlab = "Age",
     ylab = "Force of mortality", main = expression(paste("(b) ", gamma, " = 0")))
axis(1, at = c(0, 5, 10), labels = c(110, 115, 120))
axis(2, las = 1)
box()
abline(h = 0)
gamma = 1
y3 <- h(x)
plot(x, y3, type = "l", ylim = c(0, max(y3)), col = "blue", axes = FALSE, xlab = "Age", 
     ylab = "Force of mortality", main = expression(paste("(c) ", gamma, " = 1")))
axis(1, at = c(0, 5, 10), labels = c(110, 115, 120))
axis(2, las = 1)
box()
abline(h = 0)
par(oldpar)
```

\scriptsize

(a) "Life is short"; (b) "Life is unlimited but short"; (c) "Life is unlimited".

## Length-biased sampling

\scriptsize

Data are *left-truncated* and *right-truncated*:

```{r trunc, fig.width = 9, fig.height = 4}
plot(c(1999, 1999 + 5), c(110, 110 + 5), type = "l", col = "blue", lwd = 1.5, 
     xlim = c(1990, 2010), ylim = c(110, 118), las = 1, xlab = "Date", ylab = "Age - 110",
     axes = FALSE)
axis(1, at = c(1993, 1995, 1999, 2003, 2005))
axis(2, at = c(110, 114, 115, 118), las = 1, labels = c(0, 4, 5, 8))
box()
lines(c(1981, 1997), c(114, 114), lty = 3)
abline(v = c(1995, 2005), lwd = 2)
points(2004.25, 115.25, pch = "+", cex = 2)
lines(c(1993, 1995), c(110, 112), lty = 2)
lines(c(1983, 2004), c(115, 115), lty = 3)
lines(c(1995, 1997), c(112, 114), col = "blue", lwd = 1.5)
points(1997.25, 114.25, pch = "+", cex = 2)
abline(h = 110)
lines(c(2003, 2006), c(110, 113), lty = 2)
points(2006.25, 113.25, pch = "+")
```

\begin{equation*}
{\cal L} = \frac{{\cal P}(X_1 = 4)}{{\cal P}(2 < X_1 \le 12)}\times \frac{{\cal P}(X_2 = 5)}{{\cal P}(X_2 \le 6)}
\end{equation*}

*Inverse Probability Weighting* (Horwitz-Thomson, *JASA* 1952.)

## Critique of Dong et. al. by Rootzen & Zholud

\small

*   The apparent trend break in the figure is
an artifact caused by \textcolor{red}{inappropriate combination of data} from different time periods.

*   The conclusion “our data strongly suggest that the
duration of life is limited” is based on \textcolor{red}{wrong and misleading analysis}.

## Results, Rootzen & Zholud

*   "Life is unlimited but short": \textcolor{red}{$\gamma = 0$}, $\sigma = 1.34$
    *    \alert{No aging} after age 105.
    *    Expected remaining life at age $t, t > 105$: \textcolor{red}{1.34 years}.
    *    Probability of surviving one year: \textcolor{red}{47 per cent}.
*   *Interpretation:* At each birthday after 105, flip a coin. If heads, survive until next birthday, if tails, die before next birthdate.
*   \textcolor{red}{No difference} between countries, sexes, birth cohorts.



## James Vaupel, later

\Large
"I was \textcolor{red}{outraged} that \textcolor{red}{Nature}, a journal I highly respect, would publish such a \textcolor{red}{travesty}." 

## How could this happen?

\textcolor{red}{Not} uncommon:

\begin{itemize}
\item \alert{No statisticians} involved.
\begin{itemize}
\item \alert{Demographic research} is performed by people from sociology, psychology, economy, etc. with \alert{inadequate statistical training}.
\item the same holds for \alert{editors of journals} from the same fields.
\end{itemize}
\item Too \alert{powerful} statistical \alert{software}.
\begin{itemize}
\item "Who needs a statistician?"
\item "I have read the manual."
\end{itemize}
\item Failure to understand the \alert{GIGO principle} ("Garbage In, Garbage Out").
\begin{itemize}
\item There is always \alert{some} Garbage Out.
\end{itemize}
\end{itemize}

See [http://capa.ddb.umu.se/ds](http://capa.ddb.umu.se/ds)

## Data from FOB and Lisa

\scriptsize

**Birth** information (11 566 545 individuals):

```{r birthinfo,results='asis'}
soc <- readRDS("../Data/U18006_RDS/inddata.rds")
knitr::kable(head(soc, 3))
```


**Death** information (4 772 378 individuals):

```{r deathinfo, results='asis'}
dors <- readRDS("../Data/U18006_RDS/dors.rds")
knitr::kable(head(dors[, c("LINNEID", "DODSDAT")], 3))
```

**Presence** information?

## Presence information

```{r presence, fig.width = 11, fig.height = 5}
par(mfrow = c(1, 1))
fobyears = seq(1960, 1990, by = 5)
lisayears <- seq(1986, 2013, by = 1)
plot(1, 1, xlim = c(1960, 2014), ylim = c(90, 115), type = "n", axes = FALSE, xlab = "Date", ylab = "Age")
cuts <- c(fobyears + 10/12, seq(1995, 2005, by = 5), 2014)
labs <- c(paste(fobyears, "-11-01", sep = ""), paste(c(seq(1994, 2004, by = 5), 2013), "-12-31", sep = ""))
axis(1, at = cuts, labels = labs)
axis(2, las = 1, at = c(90, 95, 100, 105, 107, 109, 111, 113, 115))
box()

text(1980, 112, "dead", cex = 1.5)
abline(v = fobyears + 10 / 12, col = "blue")
abline(v = lisayears + 1, col = "red")
lines(c(1970 + 10/12 - 1, 1977 + 0.2), c(105 - 1, 105 + 7 - 10/12 + 0.2) + 1, lwd = 2)
lines(c(1970 + 10/12 - 1 - 10 + 1, 1970 + 10/12 - 1), c(105 - 10 + 1, 105), lty = 2)
points(1960 + 10/12, 96, cex = 2)
text(1965, 96, "first seen", cex = 1.5)
text(2010, 107, "last seen", cex = 1.5)
points(1977 + 0.3, 113.3 - 10/12, pch = "+", cex = 3)
lines(c(2012, 2014), c(105, 107), lwd = 2)
lines(c(2012 - 16, 2012), c(105 - 16, 105), lty = 2)
points(2014, 107, cex = 2)
abline(h = 105)
```

## Swedish data, 110+

```{r sweddat2}
s105 <- readRDS("mydata/per_105.rds")
s105$sex = factor(s105$KON, labels = c("man", "woman"))
s105$sex <- relevel(s105$sex, ref = "woman")
s110 <- eha::age.window(s105, c(110, 120))
s110 <- eha::cal.window(s110, c(1969, 2014))
source("R/drawLexis.R")
n.m <- sum(s110$KON == 1)
n.w <- sum(s110$KON == 2)
```

```{r lexis110, fig.width = 11, fig.height = 5}
par(las = 1)
with(s110[s110$KON == 2, ], drawLexis(enter, exit, event, birthdate, col = "red",
                                      at = c(1969, 1980, 1990, 2000, 2014),
                                      xlim = c(1969, 2015), ylim = c(110, 115)))
with(s110[s110$KON == 1, ], drawLexis(enter, exit, event, birthdate, col = "blue", add = TRUE))
abline(h = 110)
abline(v = c(1969, 2014), lty = 3, lwd = 2)
abline(h = c(111, 112, 113, 114), lty = 3)
legend("topleft", legend = c("Men", "Women"), col = c("blue", "red"), lty = 1)
legend("top", legend = c("Dead", "Censored"), pch = c("+", "o"))
```

\scriptsize

There are `r n.w` women and `r n.m` men.

## The Gompertz distribution

Hazard function:

\begin{equation*}
h(t; \alpha, r) = \alpha \exp(r t), \quad t > 0
\end{equation*}

If $r = 0$:

\begin{equation*}
\begin{split}
h(t; \alpha, 0) &= \alpha, \quad t > 0 \\
H(t; \alpha, 0) &= \alpha t, \quad t > 0.
\end{split}
\end{equation*}

## Cumulative hazards beyond age 110

```{r cumhaz110}
library(eha)
source("R/Hgomp.R")
##with(s110, plot(Surv(enter - 110, exit - 110, event), mark.time = FALSE, fun = "cumhaz"))
fit.c <- coxreg(Surv(enter-110, exit-110, event) ~ 1, data = s110)
fit.g <- phreg(Surv(enter-110, exit-110, event) ~ 1, data = s110, dist = "gompertz", param = "rate")
par(lwd = 2)
plot(fit.c, axes = FALSE, xlab = "Age", ylab = "Cum. hazards", lty = 2, main = "Gompertz distribution")
tic <- 0:5
axis(1, at = tic, labels = tic + 110)
axis(2, las = 1)
axis(4, las = 1)
box()
x <- seq(0, 5, length = 100)

y <- Hgomp(x, fit.g$coefficients)
lines(x, y, col = "blue")
##check.dist(fit.c, fit.g, col = c("red", "blue"), xlab = "Age")
```

r = -0.064 (0.239), mean =  `r round(exp(-fit.g$coefficients[2]), 3)`  


<!--
## Analysis, all data, 110+

\scriptsize
```{r swedanal, results= 'asis'}
library(eha)
fit <- coxreg(Surv(enter, exit, event) ~ sex + I(birthdate - 1900), data = s110)
dr = drop1(fit, test = "Chisq")
ltx(fit, dr = dr)
```

## Analysis, no censored observations, 110+

\scriptsize
```{r swedanal2, results= 'asis'}
library(eha)
fit <- coxreg(Surv(enter, exit, event) ~ sex + I(birthdate - 1900), data = s110[s110$event == 1, ])
dr = drop1(fit, test = "Chisq")
ltx(fit, dr = dr)
```

-->

## Sweden, 105+

```{r lexis105}
par(las = 1)
s105 <- eha::cal.window(s105, c(1969, 2014))
with(s105[s105$KON == 2, ], drawLexis(enter, exit, event, birthdate, col = "red" , xlim = c(1969, 2015), ylim = c(105, 115), at = c(1969, 1980, 1990, 2000, 2014)))
with(s105[s105$KON == 1, ], drawLexis(enter, exit, event, birthdate, col = "blue", add = TRUE))
abline(h = 105)
abline(v = c(1969, 2014), lty = 3)
abline(h = c(106, 107, 108, 109, 110, 111, 112, 113, 114), lty = 3)
legend("topleft", legend = c("Men", "Women"), col = c("blue", "red"), lty = 1)
legend("top", legend = c("Dead", "Censored"), pch = c("+", "o"))
```

## Analysis,  105+

\scriptsize
```{r swedanal105, results= 'asis'}
library(eha)
fit <- phreg(Surv(enter-105, exit-105, event) ~ sex + I(birthdate - 1900), data = s105, dist = "gompertz", param = "rate")
dr = drop1(fit, test = "Chisq")
ltx(fit, dr = dr)
```

Constant hazard = 0.605.

Mean = 1/0.605 = `r round(1/0.605, 2)` years.


## Baseline hazard, 105+

\scriptsize
```{r swedanal105plot, fig.width = 11}
library(eha)
fit.c <- coxreg(Surv(enter - 105, exit - 105, event) ~ sex + I(birthdate - 1900), data = s105)
##ss <- cal.window(s105, c(1969, 2014))
##fit <- coxreg(Surv(enter - 105, exit - 105, event) ~ strata(sex) + I(birthdate - 1900), data = s105)
plot(fit.c, axes = FALSE, xlab = "Age", ylab = "Cumulative hazards", lty = 2, main = "Gompertz distribution", lwd = 2)
tic <- 0:10
axis(1, at = tic, labels = tic + 105)
axis(2, las = 1)
axis(4, las = 1)
box()
x <- seq(0, 10, length = 100)

y <- Hgomp(x, fit$coefficients[3:4])
lines(x, y, col = "blue", lwd = 2)
```

$r = -0.014\; (0.026)$ 

## In perspective, 60+

```{r persp, cache = TRUE}
s60 <- readRDS("mydata/per_60.rds")
s60$sex <- s60$KON
s60 <- s60[s60$enter < s60$exit, ]
s60$enter <- s60$enter - 60
s60$exit <- s60$exit - 60
s60 <- s60[s60$sex == 2, ]
ct <- seq(1, 45, by = 1)
oj <- piecewise(s60$enter, s60$exit, s60$event, ct)
x <- seq(0, 55, length = 1000)
y <- hpch(x, cuts = ct, levels = oj$intensity)
##plot(x, y, type = "l")
n <- NROW(s60)
ss <- s60[sample(n, size = 50000, replace = FALSE), ]
fit.c <- coxph(Surv(enter, exit, event) ~ 1, data = s60)
sf <- survfit(fit.c)
fit.g <- phreg(Surv(enter, exit, event) ~ 1, dist = "gompertz", param = "rate", data = ss)
```

```{r plotpersp}
#par(lwd = 1.5)
plot(sf, fun = "cumhaz", axes = FALSE, xlab = "Age", conf.int = FALSE, ylab = "Cumulative hazards", lty = 2, 
     main = "Gompertz distribution", lwd = 2)
ct = c(0, 10, 20, 30, 37, 45, 53)
axis(1, at = ct, labels = ct + 60)
axis(2, las = 1)
axis(4, las = 1)
box()
x <- seq(0, 50, length = 5000)
y <- Hgomp(x, fit.g$coefficients)
lines(x, y, col = "blue", lwd = 2)
abline(h = 0)
abline(v = c(37, 45), lty = 3)
```

r = `r round(fit.g$coefficients[1], 3)`, exp(r) = `r round(exp(fit.g$coefficients[1]), 3)`.

<!--

## Graphical check of constant hazards, 105+

```{r grchb, fig.width = 11, eval = FALSE}
plot(fit, col = c("blue", "red"), lwd = 1.5)
fit.w <- phreg(Surv(enter - 105, exit - 105, event) ~ I(birthdate - 1900), data = ss[ss$sex == "woman", ], dist = "pch")
fit.m <- phreg(Surv(enter - 105, exit - 105, event) ~ I(birthdate - 1900), data = ss[ss$sex == "man", ], dist = "pch")
x <- seq(0, 8, length = 100)
y <- x * fit.w$hazards[1, 1]
lines(x, y, col = "darkred", lwd = 1.5)
x <- seq(0, 6.2, length = 100)
y <- x * fit.m$hazards[1, 1]
lines(x, y, col = "darkblue", lwd = 1.5)
##abline(0, fit.w$hazards[1, 1])
#abline(0, fit.m$hazards[1, 1])
```


## Women, graphical check of constant hazard

```{r graphcheck, fig.width = 11, eval = FALSE}
par(lwd = 2)
fit.e <- phreg(Surv(enter - 105, exit - 105, event) ~ I(birthdate - 1900), data = ss[ss$sex == "woman", ], shape = 1)
fit.c <- coxreg(Surv(enter - 105, exit - 105, event) ~ I(birthdate - 1900), data = ss[ss$sex == "woman", ])
check.dist(fit.e, fit.c, main = "Exponential check", col = c("red", "darkred"))
```

## Formal test of exponentiality within Weibull family

\scriptsize

```{r formtestwei, results = 'asis', eval = FALSE}
fit.w <- phreg(Surv(enter - 105, exit - 105, event) ~ I(birthdate - 1900), data = ss[ss$sex == "woman", ])
dr <- drop1(fit.w, test = "Chisq")
ltx(fit.w, dr = dr)
```

Test of **shape = 1**: *Not* rejected (in the family of Weibull distributions).

## Formal test of exponentiality within Gompertz family

\scriptsize

```{r formtestgomp, results = 'asis', eval = FALSE}
fit.g <- phreg(Surv(enter - 105, exit - 105, event) ~ I(birthdate - 1900), data = ss[ss$sex == "woman", ], dist = "gompertz", param = "rate")
dr <- drop1(fit.g, test = "Chisq")
ltx(fit.g, dr = dr)
```

Test of **rate = 1**: *Not* rejected (in the family of Gompertz distributions).

-->

<!--
## Weibull vs. Gompertz

```{r weigom, fig.width = 11,eval=FALSE}
oldpar <- par(mfrow = c(1, 2))
check.dist(fit.w, fit.c)
check.dist(fit.g, fit.c)
par(oldpar)
```
-->

## Italy, 105+ 2009--2015 (Science June 29, 2018)

\includegraphics[width=\textwidth]{myfigs/plateau.png}

## Italy results

```{r italydata, results='asis'}
italy <- readRDS("italy.rds")
levels(italy$sex) <- c("woman", "man")
##relevel(italy$sex, ref = "man")
fit <- phreg(Surv(enter, exit, event) ~ I(cohort - 1900) + sex, data = italy, dist = "gompertz", param = "rate")
dr <- drop1(fit, test = "Chisq")
ltx(fit, dr = dr)
```

## Sweden and Italy

```{r swedit, results='asis'}
it <- italy[, c("enter", "exit", "event", "sex", "cohort")]
it$country <- "Italy"
sw <- s105[, c("enter", "exit", "event", "sex", "FODELSEAR")]
sw$cohort <- floor(sw$FODELSEAR)
sw$FODELSEAR <- NULL
sw$country <- "Sweden"
sw$enter <- sw$enter - 105
sw$exit <- sw$exit - 105
swit <- rbind(sw, it)
swit$country <- factor(swit$country)
fit <- phreg(Surv(enter, exit, event) ~ strata(country) + I(cohort - 1900) + sex, data = swit, dist = "gompertz")
##dr = drop1(fit, test = "Chisq")
##ltx(fit, dr = dr)
plot(fit, fn = "sur", col = c("red", "blue"), axes = FALSE, xlab = "Age", 
     lwd = 1.5, printLegend = "topright")
ctt <- seq(0, 10, by = 2)
axis(1, at = ctt, labels = ctt + 105)
axis(2, las = 1)
box()
```

## Skellefteå-Umeå data, birth cohort 1810-1850, ages 60+

```{r skum, fig.width = 11}
library(skum)
o60 <- obs[obs$birthdate < 1850 & obs$birthdate > 1810 & obs$sluttyp != 0, ]
o60$event <- o60$sluttyp == 2
o60 <- age.window(o60, c(60, 120))
o60$event <- o60$sluttyp == 2
o60 <- o60[, c("id", "sex", "region", "birthdate", "enter", "exit", "event", "sluttyp")]
indx <- tapply(o60$id, o60$id)
o60$first <- tapply(o60$enter, o60$id, min)[indx]
o60$last <- tapply(o60$exit, o60$id, max)[indx]
o60$what <- tapply(o60$event, o60$id, max)[indx]
o60 <- o60[order(o60$id, -o60$event, -o60$exit), ]
o60 <- o60[!duplicated(o60$id), ]
o60$event <- o60$what
o60$enter <- o60$first
o60$exit <- o60$last
o60$first <- o60$last <- o60$what <- NULL
o60 <- o60[o60$enter < o60$exit, ]
fit <- coxreg(Surv(enter - 60, exit - 60, event) ~ region + sex + I(birthdate - 1840), data = o60)
fit.g <- phreg(Surv(enter - 60, exit - 60, event) ~ region + sex + I(birthdate - 1840), data = o60, dist = "gompertz", param = "rate")    
oldpar <- par(lwd = 2, las = 1)
##check.dist(fit, fit.g, col = c("red", "blue"))
plot(fit, xlab = "Age", ylab = "Cumulative hazards", 
     axes = FALSE, lty = 2, ylim = c(0, 12))
ctt <- seq(0, 50, by = 10)
axis(1, at = ctt, labels = ctt + 60)
axis(2, las = 1)
axis(4, las = 1)
box()
par(oldpar)
abline(h = 0)
abline(v = 0)
abline(v = c(10, 20, 30, 40), lty = 3)
x <- seq(0, 50, length = 1000)
y <- Hgomp(x, fit.g$coefficients[4:5])
lines(x, y, col = "blue", lwd = 2)
```

## Conclusion

\begin{itemize}
\item No evidence of limits to human life
\begin{itemize}
\item Not rejecting a null hypothesis means no information
\end{itemize}
\item The mortality plateau after age 105 seems universal.
\item Be aware of risks of sampling bias.
\item Ditto of informative censoring.
\item \textcolor{red}{The aging process}
\begin{itemize}
\item \textcolor{red}{begins at birth}, and
\item \textcolor{red}{ends at age 105}
\end{itemize}
\end{itemize}

## Extra material

## Expectation of extremes

```{r}
sur <- function(x, n = 1) 1 - (1 - exp(-x / 1.34))^n
Esur <- function(m){
    if (m <= 0){
        res <- 0 #integrate(sur, 0, Inf, n = 1)$value
    }else{
        res <- integrate(sur, 0, Inf, n = m)$value
    }
    res
}

nn <- 1:100
yval <- numeric(100)
for (i in nn){
    yval[i] <- Esur(i)
}
plot(nn, yval, ylim = c(0, 7), xlab = "No. of observations", ylab = "E(max(obs))", type = "b", col = "blue", )
##print(deaths)
```

## The original plot, expectation added

```{r orig, fig.width = 11, fig.height = 8, eval = TRUE}
par(cex = 2)
deaths <- with(all, table(year))
yl <- numeric(length(deaths))
for (i in 1:length(deaths)){
    if (deaths[i] == 0){
        yl[i] <- 110
    }else{
        yl[i] <- Esur(deaths[i]) + 110
    }
}
plot(names(deaths), yl, type = "l", col = "black", ylim = c(110, 123), lwd = 3, xlab = "Year", 
     ylab = "Max(Age at death)", lty = 1, axes = FALSE)
axis(1, at = c(1968, 1980, 1995, 2006))
axis(2, las = 1)
box()
y1 <- year[1:(1996-1968)] * 0.1531 - 191.7

maxage <- with(all, tapply(age, year, max))
points(names(maxage), maxage, col = "red", lwd = 2)
points(names(maxage)[1:21], maxage[1:21], col = "blue",lwd = 2)
x <- c(1968, 1994)
y <- 0.1531 * x - 191.07
lines(x, y, col = "blue", lwd = 3)
x <- c(1995, 2006)
y <- 667.5 - 0.2762 * x
lines(x, y, col = "red", lwd = 3)
```

## Analysis, no censored observations, 105+

\scriptsize
```{r swedanal1052, results= 'asis'}
library(eha)
fit <- phreg(Surv(enter-105, exit-105, event) ~ sex + I(birthdate - 1900), data = s105[s105$event == 1, ])
dr = drop1(fit, test = "Chisq")
ltx(fit, dr = dr)
```


## Cohort version, 105+

```{r cohortversion}
lastyr <- 2004
c105 <- s105[s105$birthdate + 105 > 1969 & s105$birthdate + 105 < lastyr, ]
n.m <- sum(c105$KON == 1)
n.w <- sum(c105$KON == 2)
par(las = 1)
with(c105[c105$KON == 2, ], drawLexis(enter, exit, event, birthdate, col = "red" , xlim = c(1969, 2015), 
                                      ylim = c(105, 115), at = c(1969, 1980, 1990, lastyr)))
with(c105[c105$KON == 1, ], drawLexis(enter, exit, event, birthdate, col = "blue", add = TRUE))
abline(h = 105)
##abline(v = c(1969, 2014), lty = 3)
abline(h = c(106, 107, 108, 109, 110, 111, 112, 113, 114), lty = 3)
legend("topleft", legend = c("Men", "Women"), col = c("blue", "red"), lty = 1)
legend("top", legend = c("Dead", "Censored"), pch = c("+", "o"))
lines(c(1969, 1969 + 8), c(105, 105 + 8), lwd = 2)
lines(c(lastyr+0.1, lastyr + 0.1 + 8), c(105, 105 + 8), lwd = 2)
```

\scriptsize

`r n.w` women, `r n.m` men.

## Cohort Analysis

\scriptsize

```{r cohanal, results='asis'}
c105$cohort <- cut(c105$birthdate + 105, c(1969, 1985, 1995, 2004))
fit <- phreg(Surv(enter, exit, event) ~ cohort + sex, data = c105, dist = "gompertz", param = "rate")
dr <- drop1(fit, test = "Chisq")
ltx(fit, dr = dr)
```


## Cumulative hazards by period

```{r bysex}
fit <- coxreg(Surv(enter, exit, event) ~ strata(cohort), data = c105)
plot(fit, col = c("blue", "red", "black"), lwd = 1.5, lty = c(1, 2, 4), xlab = "Age", fn = "surv")
abline(h = 0, v = 105)
```
